Gap formation and stability in non-isothermal protoplanetary discs

### 2014-09-22 paper/model.tex
results section edited by mkl

\subsection{Disc model and initial condition}
The disc occupies $r\in[r_\mathrm{in}, r_\mathrm{out}]$ with
$r_\mathrm{out}=25\mathrm{in}$ {\bf check: if all simulations use the
  same domain, state it here} and $\phi\in[0,2\pi]$. The initial disc is axisymmetric with
surface density profile

>>>>

\subsection{Disc model and initial condition}
The disc occupies $r\in[r_\mathrm{in}, r_\mathrm{out}]$ and
$\phi\in[0,2\pi]$. The initial disc is axisymmetric with  
surface density profile

### 2014-09-22 paper/results.tex
results section edited by mkl

Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
and we consider 
%for a time up to $200P_0$ allowing high resolution and detection of high
%$m$ structures.
cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
and slowly cooled discs.

@@

We find very similar gap profiles and mode growth rates for
$\tilde{\beta}=100$ as with $\tilde{\beta}=10$. The disc heats up to
values $h\simeq0.06$ in the nearly adiabatic case. This is close to
the original temperature of $h=0.05$, so growth rates are not expected
to change significantly according to the linear theory presented in
\citep{li00}.

@@

{\bf
If there's time, repeat planet off simulations but keep the planet on
until $t=40P_0$ or $50P_0$. Look at stability properties when the gaps
are generally deeper (including the nearly adiabatic case). Although
with the planet instability may already appear by $t=40P_0$, the
azimtuhal average performed prior to perturbations will erase it. 
}

@@

Long term simulation of edge vortices were done for
$\tilde{\beta}=0.1,0.5,1.0,5.0,10.0$ up to a total of
$2000P_0$. Planets were left free to interact dynamiclly with the disk
and vortices after $t=30P_0$ as apposed to previous section. These
simulations were less resolved with $(N_r,N_{\phi})=(512,1024)$ in
order for them to be computationally economical. An open boundry
condition was used in the outer edge of
$r_{\mathrm{out}}=45r_{\mathrm{in}}$ allowing significant room for
linblad resonances of the vortices.

After the semi-linear growth of vortices, vortex merger effects take
hold and by $150P_0$ for all $\tilde\beta$ values, only $m=1$ vortex
modes exist. Plots of the amplitude of the $m=1$ mode for the
different $\tilde\beta$ cases are shown in
Fig~\ref{lifetimeplot}. These instabilities stay in quasi-steady
states for $>1000P_0$ in most cases, in which they grow in intensity
and reaching overdenisities almost 10 times the intial local
density. The growth of these vortices are supplied by the continuous
generation of vorticity by the planet-disk interactions. Eventually
the $m=1$ amplitudes are also characterized by sudden large drops of
vortex intensity which happen over timescales of $50P_0$. After these
drops the instability is never seen to reform at such large over
densities and density bumps are never larger than the planet induced
shocks. We designate the time of these amplitude drops as the lifetime
of the quasi-steady vortex.

Rossby values have dynamics that are coupled with the lifetime and
$m=1$ amplitude of the vortices. As the vortex forms it has a
charcteristic value of $Ro\approx-0.1$ indicating anti-cyclonic
motion. As the $m=1$ amplitude grows the relative spin of the vortex
continuously grows along with it, reaching values of
$Ro\approx-0.4$. As the vortex dies off as described above the
associated Rossby number of the vortex accordingly quickly approaches
0.

The sudden dissipation of these vortices seem to be a result of the
vortices shocking the system. Measurements of the Mach number near the
vortices show large increases in value when the vortex is dissipating
which indicates that the system is more susceptible to shocks. Also
gradients of surface density are seen to spike up in wake like
features around the vortex during dissipation which is also indicative
of shocks. Of note is also that the large overdensity the vortex
creates during it's quasi-steady state distorts the background
disk. So as the vortex grows in intensity it alters the intial gap
structure that formulated its growth.

Contradictory to our `planet-off' simulations, vortex lifetimes did
not monotonicly depend on intial edge stability and from our
simulations we find that there is $t_{\mathrm{coo}l}$ which optimizes
the vortex lifetime as seen in Fig~\ref{lifetimeplot}. Increasing
$\tilde\beta$ correspondingly increased the lifetime of the vortex up
to a critical value for $\tilde{\beta}=5.0$ which had a lifetime that
lasted beyond the simulation time of $2000P_0$. Cooling rates with
$\tilde\beta>5.0$ showed lifetimes that were well below the max
lifetime. By analyzing the aspect ratio of the corotation region
around the vortex we find that longest lifetime $\tilde\beta=5.0$ disk
corresponds to $h\approx0.06$ which is a result similarily found by
\citet{fu14} to be the optimum lifetime for long-term simulations of
isothermal disks with varying intial temperature profiles $h$.

Competing effects of gap stability with $t_{\mathrm{coo}l}$ result in
the the non-montonic form of the lifetime. As the $\tilde\beta$
increases so does the stability of the gap edge as seen in
section~\ref{linear}, however as $\tilde\beta$ increases so does the
$c_{\mathrm{iso}}$ which is known from theory to effect the growth of
instability \citep{li00}. Thus increasing $\tilde\beta$ increases the
sound speed $c_{\mathrm{iso}}$ and likewise the gap instability up to
a certain point in which the effects of the more stable gap edge
becomes a significant effect. The correlation of growth rate and sound
speed is non-negligible as in previous `planet-off' case as the disk
now can reach values of $h\simeq0.08$ for very slow cooling and the
importance of the effect can be imagined to accumulate over the now
considerably longer timescale. These effects dictate when then the
vortex reaches the critical value in which it shocks the system that
subsequently causes disspation and hence the lifetime fo the vortex.

@@

\begin{figure}
  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
  \caption{Cartesian plot of relative density pertibation for
    $\tilde\beta=0.1$ vortex case during the quasi-steady state. The
    large scale non-axisyemetric $m=1$ overdensity can be
    seen. \label{Vortex2D}} 
\end{figure}

%\begin{figure}
%   \includegraphics[scale=.42]{figures/stability_vis6betalow.ps}
%   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=0.1$. }
% \label{stability_vis6lowb)}
% \end{figure}

%\begin{figure}
%   \includegraphics[scale=.42]{figures/stability_vis6betamed.ps}
%   \caption{Same as Fig. \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=1.0$. }
% \label{stability_vis6medb)}
% \end{figure}
%Same as \ref{stability_vis9lowb} but $\hat{\nu}=10^{-6}$ and $\tilde{\beta}=1.0$.

%we checked that open bc doesn't affect gap structure
%checked that t=200 plots are similar to t=100

%surface density plot

%aspect-ratio plot

%general vortensity plot

%% \begin{figure}
%%   \includegraphics[scale=.42,clip=true,trim=0cm 1.8cm 0cm 0cm]{figures/compare_profiles_dens020.ps}\\
%%   \includegraphics[scale=.42]{figures/compare_profiles_h020.ps}
%%   \caption{Steady-state gap profiles in a low mass viscous disc. The
%%     surface density perturbation (top) and disc aspect-ratio (bottom)
%%     are shown as a function of the cooling parameter:  
%%     $\tbeta=0.1$ (solid, fast cooling), $\tbeta=1$ (dotted,
%%     moderate cooling) and $\tbeta=10$ (dashed, slow
%%     cooling). \label{lvisc_steady_gap}}  
%% \end{figure}

%% \begin{figure}
%%   \includegraphics[scale=.42]{figures/compare_profiles_gvort020.ps}
%%   \caption{Gap structure in a low mass viscous disc, in terms of the
%%     perturbed generalized vortensity as a function of the cooling
%%     parameter. \label{lvisc_steady_gvort}} 
%% \end{figure}

%%\section{Gaps in massive discs (MKL)}
%% We first examine disc models with $Q_o=1.5$. We set the physical
%% viscosity $\hat{\nu}=10^{-9}$, so that the only energy source is
%% through shock-heating (via artificial viscosity) and the $\mathcal{C}$
%% function when $e\Sigma<e_i\Sigma_i$. The numerical resolution is
%% $N_r\times N_\phi = 512\times 1024$.

%% %Since we are primarily concerened with gap stability, it is important
%% %to first examine the gap structured opened by the planet as a function
%% %of cooling. 
%% Fig. \ref{gvort1d_q1d5} compares the gap structure in terms of the GV profile for three
%% levels of cooling: $\tilde{\beta}=0.1$, $\tilde{\beta}=1$ and
%% $\tbeta=10$. For convenience we will refer to these cases as fast,
%% intermediate and slow cooling, respectively. The snapshot is taken at
%% $t=30P_0$, just after the planet is fully introduced.

%% In terms of the GV profile, the qualitative features of the planetary
%% gap --- localized GV extrema at the gap edges --- remain unchanged
%% despite two orders of magnitude difference in the cooling rate.

%% \begin{figure}
%% %  \includegraphics[scale=.42,clip=true,trim=0cm 1.8cm 0cm 0cm]{figures/dens1d_q1d5_003_global.ps}\\
%% %  \includegraphics[scale=.42]{figures/gvort1d_q1d5_003_global.ps}
%%   \caption{Gap structure in terms of the perturbed surface density
%%     (top) and perturbed generalized
%%     vortensity (bottom), as a function of the cooling parameter:  
%%     $\tbeta=0.1$ (solid, fast cooling), $\tbeta=1$ (dotted,
%%     intermediate cooling) and $\tbeta=10$ (dashed, slow
%%     cooling). \label{gvort1d_q1d5}} 
%% \end{figure}

%% \begin{figure*}
%% %  \includegraphics[scale=.55,clip=true,trim=0cm 0cm 0cm
%% %    0cm]{figures/noniso0_HR_dens004}\includegraphics[scale=.55,clip=true,trim=2.26cm 0cm 0cm
%% %    0cm]{figures/noniso1_HR_dens004}\includegraphics[scale=.55,clip=true,trim=2.26cm 0cm 0cm
%% %    0cm]{figures/noniso2_HR_dens004}
%%   \caption{Gap instability in the heavy disc model, as a function of
%%     the cooling parameter: $\tbeta=0.1$ (left), $\tbeta=1$ (middle)
%%     and $\tbeta=10$ (right). The relative surface density
%%     perturbation is shown. \label{polarxy_q1d5}} 
%% \end{figure*}

>>>>

Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
with $r_\mathrm{out}=25r_\mathrm{in}$, 
and we consider 
%for a time up to $200P_0$ allowing high resolution and detection of high
%$m$ structures.
cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
and slowly cooled discs.  {\bf boundary condition for these sims?}

@@

We find very similar gap profiles and mode growth rates for
$\tilde{\beta}=100$ as with $\tilde{\beta}=10$. The disc heats up to
values $h\simeq0.06$ in the nearly adiabatic case. This is close to
the original temperature of $h=0.05$, so linear growth rates are not expected
to change significantly %according to the linear theory presented in
\citep{li00}.

@@

{\bf
If there's time, do low resolution planet off simulations but keep the planet on
until $t=40P_0$ or $50P_0$. Look at stability properties when the gaps
are generally deeper (including the nearly adiabatic case). Although
with the planet instability may already appear by $t=40P_0$, the
azimtuhal average performed prior to perturbations will erase it. 
}

@@

We now consider long term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
throughout.  
% up to a total of
%$2000P_0$. 
%Planets were left free to interact dynamiclly with the disk
%and vortices after $t=30P_0$ as apposed to previous section.  
We employ a grid with $(N_r,N_{\phi})=(512,1024)$ 
in order for these simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
effects on vortex evolution. We apply an open boundary at
$r_\mathrm{out}$. {\bf mention lower resolution preliminary runs -
  similar results to those presented here (?). want to demonstrate convergence}

%An open boundry
%condition was used in the outer edge of
% allowing significant room for
%linblad resonances of the vortices.

\subsection{Generic evolution}

The linear growth of the RWI and vortex-formation is followed by 
vortex merging. For all $\tilde\beta$ values, only one vortex remains
by $150P_0$.
%i.e. only the $m=1$ mode remains. 
{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
  taken to reach one vortex as a function of cooling}
%by $150P_0$ for all , only $m=1$ vortex
%modes exist. 
The evolution of the amplitude of the $m=1$ mode in surface density is
shown in Fig~\ref{lifetimeplot} for   
different $\tilde\beta$. {\bf should state (also for planet-off runs)
  what the radial range is used for averaging the Fourier amplitudes}

@@

In all cases the system remains in a quasi-steady state for
$\gtrsim500P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex grows in intensity. As the vortex forms it has a
charcteristic value of $Ro\approx-0.1$ indicating anti-cyclonic
motion. As the $m=1$ amplitude grows the relative spin of the vortex
continuously grows along with it, reaching values of
$Ro\approx-0.4$, corresponding to surface density almost 10 times the
initial value. {\bf for what cooling are these values quoted for? do
  these values depend on cooling?} Vortex growth is attributed to
the continuous generation of vorticity by the planet-disc interactions. 
{\bf is the evidence for this the fact that growth is not observed in
  planet-off simulations?}

{\bf note that there
  are two origins for the surface density boost relative to $t=0$: 1) gap
  formation, which increases the surface density at the location of
  the instability (since they happen at surface density max., and 2)
  vortex instability). what is the surface density boost relative to
  the gap profile?}

Interestingly, in most cases we observe a sudden drop in 
the $m=1$ amplitude within the simulation time, which occurs over
timescales of $50P_0$, i.e. much shorter than the duration of the quasi-steady
state. {\bf is this dissipation timescale independent of cooling?}.  
%Eventually
%the $m=1$ amplitudes are also characterized by 
After these characteristic drops, the instability is never seen to
reform at such large overdensities and surface density bumps are never larger
than that due to the planetary wakes. We designate the time of these
amplitude drops as the lifetime 
of the quasi-steady vortex.  
{\bf question: does instability
  actually happen again? if it does, but just weaker, that may be
  because the first vortex has smoothed out the gap edge, so
  instabilities following that would be weaker.}

\begin{figure}
  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
  \caption{Relative surface density perturbation for the
    $\tilde\beta=0.1$ case during quasi-steady state with a single
    vortex. The plot for other values of the cooling time $\tilde{\beta}$ are
    similar. {\bf is this plot truncated? in the text it says the
      domain goes to $r=45$?}
    %The large scale non-axisyemetric $m=1$ overdensity can be
    %seen. 
    \label{Vortex2D}} 
\end{figure}

{\bf
\subsection{Additional analysis on vortex dissipation}
Conjecture: vortex death due to vortex shocking the system and
dissipating its energy. discuss one case. here we want to explain or
suggest why vortices die} 
%Rossby values have dynamics that are coupled with the lifetime and
%$m=1$ amplitude of the vortices.

As the vortex dies off as described above its associated Rossby number
of quickly approaches $0$. Gradients of surface density are seen to
spike up in wake-like features around the vortex during dissipation. 
{\bf show
  this. e.g. vortex before, during and after death. maybe show
  snapshots of surface density perturbation (vortex) compared with
  $\nabla\Sigma/\Sigma$ (vortex shocks). demonstrate vortex shape
  changes and shocks develop prior ot death. state whether this is
  generic for all coolings}.  
In addition, we also observed large increases in the Mach number near
the vortex during dissipation, which indicate susceptibility to
shocks. {\bf show mach number plot, or quote values. relevant mach
  number is fluid flow speed with respect to 
  vortex bulk motion around the disc. does the typical mach number
  near the vortex just before its death depend on cooling?} 
Thus, the sudden drop in vortex amplitude may  
be a result of the development of associated spiral shocks dissipating
its energy \citep[cf.][who 
  suggested damping due to planet-induced shocks]{fu14}.

{\bf note that fu et al suggested vortex damping by the planet
  shocks. here, we are suggesting vortex dissipates energy by shocking
  the disk. our evidence is that we observe vortex growth before it
  dissipates, and spiral shocks associated with the vortex always
  develop before vortex death. once it starts shocking the disc it
  loses energy. if the planet-induced shocks were responsible, death
  should be more gradual. can we think of other evidences that it's
  not the planet shock that's responsible?                    
}

Vortex growth during the quasi-steady state also acts to smooth out
the gap edge. 
%Of note is also that the large overdensity the vortex
%creates during it's quasi-steady state distorts the background
%disk. 
{\bf show this (smoothening of gap edge due to vortex growth),
  e.g. general vortensity evolution or surface density 
  evolution}. 
This is because the vortex has an associated viscosity {\bf quote a
  typical value of alpha viscosity associated with vortex}. This acts
against gap-opening by the planet, so that the condition for the RWI
becomes less favourable as the vortex grows. {\bf self-limiting
  behaviour?}

\subsection{Vortex lifetimes as a function of cooling rate}

{\bf describe dependency on
cooling. Do we have enough data to have a `lifetime v.s. beta'
  plot? here we want to explain why vortex lives longer with longer
  cooling, but very long cooling results in shorter lifetimes. 
  Consider the following quantities as a function of cooling:
  initial strength of the single vortex; growth rate of single vortex
  strength from its formation to just before death; ampiltude of
  vortex just before death; disc temperature in the vortex region.

  we expect the initial $m=1$ vortex (after merging) to be weaker with
  increasing cooling time because the instability is weaker. (see
  meheut et al 2013, mnras, 430, 1988 who show that the initial vortex
  amplitude after linear growth is correlated with the linear growth
  rate).

  suppose the vortex shocks when it reaches a certain amplitude (which
  probably increases with longer cooling rate because it's more
  difficult to shock a hotter disc). as we
  increase the cooling time, the initial $m=1$ vortex is weaker, it
  also grows more slowly because gap-opening is opposed by a hotter
  disc. the above implies that it takes longer for the vortex reach
  the shock-inducing amplitude and dissipate.

  but the above is not consistent with non-monotonic behaviour, which
  predicts the $\tilde{\beta}=10$ vortex to survive longest. this is
  because we expect the initial vortex to be weakest (check) which
  means it should take the longest to grow and shock. check
  whether there's anything special about this case. one possibility is
  that vortex growth during the quasi-steady state is actually faster
  because of the raised temperature (li et al 2000 show increased
  temperature favours RWI - possibiliy evidenced here by a stronger
  vortex before death than other cases)

  perhaps have a case of $\tilde{\beta}=7.5$ as further evidence of
  non-monotonic behaviour
}

Fig. \ref{lifetimeplot} shows that increasing $\tilde{\beta}$
increases the lifetime of the vortex up to a critical 
value $\tilde{\beta}=5.0$ for which the vortex lasted beyond the
simulation time of $2000P_0$. This is consistent with vortex
dissipation when it induce shocks. Our `planet-off' simulations
indicate weaker vortices are formed initially with increased cooling
times. After merging, the growth of the single vortex, mediated by disc-planet
interaction, is expected to be slower with increased cooling time
because gap-opening becomes more difficult in a hotter
disc. Furthermore, the increased sound-speed suggest the vortex should
reach larger amplitudes in order to induce shocks. 
These expectations
imply that it takes longer for the vortex to induce shocks and
dissipate, and hence longer lifetimes with increasing
$\tilde{\beta}$.

However, Fig. \ref{lifetimeplot} shows that increasing the cooling
time further to $\tilde{\beta}=10$ actually results in a
\emph{shorter} vortex lifetime. Notice for $\tilde{\beta}=10$ the vortex growth 
during the quasi-steady state is faster than for
$\tilde{\beta}=5$. This may be attributed to the higher temperature
reached in the former case ($h\simeq 0.08$). {\bf temperature for
  $\tilde{\beta}=5$?}, since \cite{li00} showed that the RWI grows
faster with increased disc temperature. At long cooling times, this
effect may dominate over the weaker gap-opening effect at higher $h$.

Thus, unlike our `planet-off' simulations we find that there is
a cooling time $t_c$ which maximises the vortex lifetime. {\bf to make
this comparison we need to discuss vortex lifetimes for planet-off
simulations as well} 
This
non-monotonic behaviour was also reported by \cite{fu14}, who
performed isothermal disc-planet simluations with different fixed
aspectratios.   
%Cooling rates with
%$\tilde\beta>5.0$ showed lifetimes that were well below the max
%lifetime. 
By analyzing the region
around the vortex {\bf define this region} we find the
$\tilde\beta=5.0$ case, with the longest vortex lifetime, has
$h\approx0.06$. This value of $h$ was also found to maximize vortex
lifetimes in the isothermal simulations of \cite{fu14}.

%be the optimum lifetime for long-term simulations of
%isothermal disks with varying intial temperature profiles $h$.

%Competing effects of gap stability with $t_{\mathrm{coo}l}$ result in
%the the non-montonic form of the lifetime. As the $\tilde\beta$
%increases so does the stability of the gap edge as seen in
%section~\ref{linear}, however as $\tilde\beta$ increases so does the
%$c_{\mathrm{iso}}$ which is known from theory to effect the growth of
%instability \citep{li00}. Thus increasing $\tilde\beta$ increases the
%sound speed $c_{\mathrm{iso}}$ and likewise the gap instability up to
%a certain point in which the effects of the more stable gap edge
%becomes a significant effect.

%The correlation of growth rate and sound
%speed is non-negligible as in previous `planet-off' case as the disk
%now can reach values of $h\simeq0.08$ for very slow cooling and the
%importance of the effect can be imagined to accumulate over the now
%considerably longer timescale. These effects dictate when then the
%vortex reaches the critical value in which it shocks the system that
%subsequently causes disspation and hence the lifetime fo the vortex.
